Clustering based on probability distributions with application on residential customers

### 2021-10-13 check_work.Rmd
wkndwday pattern changed to boxplot

theme_characterisation <- function() {
theme_bw() + 
      theme(strip.text = element_text(size = 7, margin = margin(b = 0, t = 0))) +  + theme_bw() +
   theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  theme(panel.spacing =unit(0, "lines")) + 
    theme(axis.text.x = element_text(angle=90, hjust=1, size = 7)) + theme(legend.position = "bottom")+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm")) +
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00"))+
  theme(legend.position = "bottom")
}
```

# Introduction

<!-- Description of data available -->

<!-- significance of load clustering -->
Large spatio-temporal data sets, both from open and administrative sources, offer up a world of possibilities for research. <!--Beyea (2010) has pointed out, there has been little exploration of the full potential of these data bases and their benefits can reach beyond the original intentions for collecting these data.--> One such data sets for Australia is the Smart Grid, Smart City (SGSC) project (2010–2014) available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee)<!--and Data61 CSIRO-->. The project provides half-hourly data of over 13,000 household electricity smart meters distributed unevenly from October 2011 to March 2014. <!---The massive amount of data generated in such projects could be overwhelming for analysis.--> Raw data of these asynchronous time series can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Electricity utilities can utilize the smart meter usage patterns to develop targeted tariffs for individual groups and alleviate the problem of volatility in production by capitalizing on the flexibility of consumers.

<!-- Sources of variation in large data sets -->
<!-- The enormous quantity of data provides for greater individual level clarity and analysis. However, -->
<!-- due to the growing variety of consumers,  -->

\noindent Larger data sets include greater uncertainty about customer behavior due to growing variety of customers. Households are distributed geographically, and have different demographic properties such as the existence of solar panels, central heating or air conditioning. Multiple temporal dependencies define the behavioral patterns, which vary from customer to customer. Some families, for example, use a dryer to dry their clothing, while others hang them to dry on a line. This may be reflected in their weekly profile. They may have monthly variations where some customers are more prone to use air conditioners or heaters than others despite the existence of comparable electrical equipment and being subjected to similar weather conditions. The variations in behavior may occur on a regular basis, with some consumers being night owls and others being morning larks. Day-off energy usage may vary depending on whether consumers choose to remain at home or engage in outdoor activities. These various causes of variations in energy behavior, ranging from age, lifestyle, and family composition to building characteristics, weather, presence of various electrical equipments and others, make the problem of effectively segmenting consumers into similar energy behavior a particularly intriguing one.

<!-- unsupervised clustering -->
\noindent This problem becomes more difficult when there is no other information on the customers besides their time series of energy usage. It is likely that such additional information is not available in order to protect the customers' confidentiality. Furthermore, it is not guaranteed that energy providers would always update customer profiles, such as property characteristics, in a timely manner whenever they change.
So this study does not attempt to explain why consumption differs. Instead, the work looks at how much energy-use heterogeneity can be uncovered in smart meter data and what are some of the most typical electricity use patterns by simply using the time series.

<!-- This work -->
This is similar to a stochastic approach (@Motlagh2019-yj) to clustering, which proposes interpreting electricity demand as a random process and extracting time-series characteristics, or a model of the series, to enable unsupervised clustering. Unsupervised clustering is only as good as the features that are extracted/selected or the distance metrics that were utilized. Well-designed additional features may collect characteristics that default features cannot. Based on the underlying structure of the temporal data, this article offers new distance metric and features for clustering and applies them to actual smart-meter data. Firstly, the distance metric is based on probability distribution, which in our knowledge is the first attempt to cluster smart meter data using probability distributions. These recorded time series are asynchronous, with varying time lengths for different houses and missing observations. Taking probability distributions helps to deal with such data, while helping with dimension reduction in one hand but not losing too much information due to aggregation. Secondly, we recognise that most clustering algorithms only provide hourly energy profiles during the day, but this approach provides a wider approach to the issue, seeking to group consumers with similar shapes over all important cyclic granularities. Since cyclic granularities are considered instead of linear granularities, clustering would group customers that have similar repetitive behavior across more than one cyclic granularities across which patterns are expected to be significant.

<!-- lit review -->
_Related work_

A typical clustering technique includes the following steps: (a) establishing distance (dissimilarity) and similarity through feature or model extraction and selection; and (b) selecting the clustering algorithm design. Distance measures could be time-domain based or frequency-domain (fast Fourier transform). A model-based clustering works by transforming the series into other other objects such as structure or set of parameters which can be more easily characterised and clustered (@Motlagh2019-yj). [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. The essential temporal characteristics of the curves are defined or extracted using feature-based clustering. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. None of the $34$ articles chosen for their emphasis use Australian smart meter data. The most often used algorithm is K-Means. Using K-Means without considering time series structure or correlation results in inefficient clusters. Principal Component Analysis (PCA) or Self-Organizing Maps (SOM) eliminate correlation patterns and decrease feature space, but lose interpretability. To reduce dimensionality, several studies use principal component analysis or factor analysis to pre-process smart-meter data before clustering (@Ndiaye2011-pf). Other algorithms utilised in the literature include k-means variants, hierarchical approaches, and greedy k-medoids. Time series data, such as smart metre data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics using Fourier transformation, which converts data from time to frequency and then uses K-Means to cluster by greatest frequency .

<!-- Most clustering algorithms offer only daily energy profiles throughout the hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar shapes across all significant cyclic granularities -->

<!-- There are two reasons why any additional consumer level data would not be available. Customer profiles, like property features, vary over time and are not always updated in a timely manner by energy suppliers. Also such data might not be available when the anonymity of the customers are of utmost importance.  -->

<!-- Segmenting unlabeled time series patterns has received little attention in the literature. This is a situation when there is no additional information on the socio-economic charactertistics, property types or family size of the customers. There are two reasons why any additional consumer level data would not be available. Customer profiles, like property features, vary over time and are not always updated in a timely manner by energy suppliers. Also such data might not be available when the anonymity of the customers are of utmost importance. So this study does not explain why consumption differs. Instead, this work investigates how much energy usage heterogeneity can be found in smart meter data and what some of the most common electricity use patterns are. -->

<!-- These recorded time series are asynchronous, with varying time lengths for different houses and missing observations. -->

<!-- An aggregation approach to reduce the data size is not exactly ideal when dealing with highly heterogeneous profiles -->

<!-- lit review and limitations if any -->
<!-- _Common load clustering techniques of smart meter data_ -->

<!-- A lot of work has been done in the literature on clustering customers by their load characeristics. Customers should be clustered uniquely, with behaviors represented with distinctive profile patterns. Approaches could be deterministic or stochastic. Deterministic approaches are hard to generalize due to the need for detailed knowledge of occupancy, appliances, and their time of use. In contrast, stochastic load analysis does not require extensive data collection other than the time series itself [13,14]. This technique -->
<!-- suggests viewing load as a random process, where extracting time-series features, or a model of the series, facilitates unsupervised clustering. Any standard clustering methodology consists of the following steps: (a) defining distance (dissimilarity) and similarity mostly through feature extraction and selection; and (b) choosing the clustering algorithm design. -->
<!-- Popular distance measures could be time-domain based or frequency-domain (fast Fourier transform). Any model based extraction works by transforming the series into other other objects such as structure or set of parameters which can be more easily characterised and clustered. A type of clustering based on information theory such as Shannon or Renyi entropy and their variants are addressed in [42].In contrast to models, a feature-based extraction is used to explicitly define or automatically extract the curves’ key time features, for instance by application of PCA on the daily curves. @Tureczek2017-pb presented a systematic review of the current state of the art in smart meter data analytics, which evaluates approximately 2100 peer-reviewed papers and summarizes the main findings. None of the $34$ selected papers which focus on clustering consumption are based on Australian smart meter data. -->
<!-- The algorithm most ubiquitously employed is K-Means. But the omission of the time series structure or correlation in the analysis while employing K-Means leads to inefficient clusters. Principal Component Analysis or Self-Organizing Maps removes correlation structures and transitions the data to a reduced feature space, but it comes at a cost of interpretability of the final results. Some papers present pre-processing of the smart-meter data before clustering through principal component analysis or factor analysis for dimensionality reduction or self-organizing maps for 2-Dimensional representation of the data (@Ndiaye2011-pf). Other algorithms used in the literature include k-means variations, hierarchical methods and k-medoids based on a greedy algorithm have been designed to select typical periods in the time series. -->

<!-- Time-domain analysis is preferred to specify time characteristics, while methods such as fast Fourier transform are used to reveal frequencies that highlight recurrent daily and annual load variations [27]. Popular time-domain techniques, such as principal component analysis -->
<!-- (PCA), detect key features of daily curves and reasonably link them to -->
<!-- distinctive demand behaviours [28,29]. For instance, PCA can help -->
<!-- reduce diurnal dimensions, e.g., 48 half-hourly power measurements, down to a few orthogonal components. @Tureczek2017-pb presented a systematic review of the current state of the art in smart meter data analytics, which evaluates approximately 2100 peer-reviewed papers and summarizes the main findings. None of the $34$ selected papers which focus on clustering consumption are based on Australian smart meter data. -->
<!-- The algorithm most ubiquitously employed is K-Means. But the omission of the time series structure or correlation in the analysis while employing K-Means leads to inefficient clusters. Principal Component Analysis or Self-Organizing Maps removes correlation structures and transitions the data to a reduced feature space, but it comes at a cost of interpretability of the final results. Some papers present pre-processing of the smart-meter data before clustering through principal component analysis or factor analysis for dimensionality reduction or self-organizing maps for 2-Dimensional representation of the data (@Ndiaye2011-pf). Other algorithms used in the literature include k-means variations, hierarchical methods and k-medoids based on a greedy algorithm have been designed to select typical periods in the time series. -->

<!-- The clustering is frequently applied directly to the raw data without scrutinizing for auto correlation and periodicity. -->

<!-- The key challenge is the application of appropriate processes to reduce the extreme dimensionality of load time series to facilitate unique clusters. Time feature extraction is a potential remedy, however, it is limited by the type of noisy, patchy, and unequal time-series common in residential datasets.  -->

<!-- The literature -->
<!-- suggests the clustering of residential customers by their load characteristics. The key challenge is the application -->
<!-- of appropriate processes to reduce the extreme dimensionality of load time series to facilitate unique clusters. -->
<!-- Time feature extraction is a potential remedy, however, it is limited by the type of noisy, patchy, and unequal -->
<!-- time-series common in residential datasets. In this paper we propose a strategy to alleviate these limitations by -->
<!-- converting any types of load time series into map models that can be readily clustered. -->

<!--Unlabelled data -->
<!-- While forecasting techniques for extremely granular temporal data are readily accessible, -->

<!-- The basic idea behind the clustering algorithm based on distribution is that the data, generated from the same distribution, belongs to the same cluster if there exists several distributions in the original data.  -->

<!-- Segmenting unlabeled time series patterns has received little attention in the literature. This is a situation when there is no additional information on the socio-economic charactertistics, property types or family size of the customers. There are two reasons why any additional consumer level data would not be available. Customer profiles, like property features, vary over time and are not always updated in a timely manner by energy suppliers. Also such data might not be available when the anonymity of the customers are of utmost importance. So this study does not explain why consumption differs. Instead, this work investigates how much energy usage heterogeneity can be found in smart meter data and what some of the most common electricity use patterns are. -->

<!-- lit review and limitations if any -->
<!-- _Common load clustering techniques of smart meter data_ -->

<!-- An aggregation approach to reduce the data size is not exactly ideal when dealing with highly heterogeneous profiles, as (XXX big data to the rescue example) demonstrates. -->
<!-- The foundation for this study is Tureczek2017-pb, which conducts a systematic review of the current state of the art in smart meter data analytics, which evaluates approximately 2100 peer-reviewed papers and summarizes the main findings. None of the 34 selected papers which focus on clustering consumption are based on Australian smart meter data. The clustering is frequently applied directly to the raw data without scrutinizing for auto correlation and periodicity. The algorithm most ubiquitously employed is K-Means. But the omission of the time series structure or correlation in the analysis while employing K-Means leads to inefficient clusters. Principal Component Analysis or Self-Organizing Maps removes correlation structures and transitions the data to a reduced feature space, but it comes at a cost of interpretability of the final results. @Tureczek2018-ha has shown that a transformation of data to incorporate autocorrelation before K-Means clustering can improve performance and enable K-Means to deliver smaller clusters with less within-cluster variance. However, it does not explain the cluster composition by combining it with external data. Some papers present pre-processing of the smart-meter data before clustering through principal component analysis or factor analysis for dimensionality reduction or self-organizing maps for 2-Dimensional representation of the data (@Ndiaye2011-pf). Other algorithms used in the literature include k-means variations, hierarchical methods and k-medoids based on a greedy algorithm have been designed to select typical periods in the time series. As the methods are often situation specific,  -->
<!-- it makes sense to compare them on the performance rather than any standard performance metric. A type of clustering based on information theory such as Shannon or Renyi entropy and their variants are addressed in , which differs from typical methods adopted for electricity consumer classification, based on the Euclidean distance notion. @Motlagh2019-yj presents strategy to address the problems on patchy, and unequal time-series common in residential data sets by converting load time series into map models. Most time-series clustering models are limited to handling time domain with same start and end date and time. Most of the solutions to handle this like longest common subsequence, dynamic time warping are prone to computational limit with increased length of the series. -->

<!-- our approach and contribution -->
<!-- probability distribution AND cyclic granularities -->

<!-- Large data sets -->
<!-- On one hand, the sheer volume of data enables greater clarity and analysis at the individual level. On the contrary, it also provides challenges as larger data sets have greater uncertainty about customer behavior due to increased variety of customers. Additionally, when using big data, there is often a temptation to aggregate the data to reduce its volume so it is easy to manage and computationally feasible to analyse but (XXX big data to the rescue example) show that such a strategy may not always be an optimal choice especially if we are to consider highly heterogeneous profiles that have a dimension of human behaviour. -->

<!-- how looking at probability and cyclic granularity helps -->

<!-- how is your work different from the rest of the literature -->

<!-- \noindent The massive amount of data generated in such projects could be overwhelming for analysis. Electricity utilities can utilize the consumption patterns of customers to develop targeted tariffs for individual groups and alleviate the problem of volatility in production by capitalizing on the flexibility of consumers. Beyea (2010) has pointed out, there has been little discussion or exploration of the full potential of these data bases and their benefits can reach beyond the original intentions for collecting these data. Thus, there is a scope to investigate and analyze these data in various ways for a greater understanding of consumption patterns and how they correlate with other economic, physical or geographical factors. In this work, we are interested to see how we can utilize this dataset to group different customers with similar periodic behavior.  -->

<!-- Towards this goal, this chapter aims to: (a) describe the contents of the data set in SGSC database that we can utilize, and (b) propose a clustering algorithm to group customers with similar periodic behaviors. The distance metric introduced in Chapter 2 will be the inputs for this cluster analysis. One of the advantages of using our  -->
<!-- approach is that the technique is based on probability distributions instead of raw data. Many clustering approaches are limited by the type of noisy, patchy, and unequal time-series common in residential data sets. Since the distance measure considered is based on differences in probability distribution of time series, it is likely to be less sensitive to missing or noisy data. -->

<!-- _Themes_ -->

<!-- - Dimension reduction: If each $P_{i, j, k}$ be considered to be a point in the space, key $i$ would have $mp$ dimensions as opposed to $n_i$ dimensions in case of considering raw data. Hence for a large number of observations ($n_i>>mp$), this approach benefits by transitioning to a lower dimension. -->

<!--  - Avoid loss of information due to aggregation: This approach ensures key characteristic information of the data is not lost due to averaging or aggregation measures in an attempt to transition to a lower dimension. Hence, this approach could be thought to somehow balance the drawback of considering raw data or aggregated data. -->

<!-- - Robustness to outliers: This approach could be adapted to be robust to outliers and extreme behaviors by trimming the tails of the probability distributions. -->

<!-- - Non-synchronized observed time periods: Considering probability distribution would imply the clustering process can handle keys that are observed over periods of time that are overlapping but don't necessarily coincide. -->

<!-- - Similar periodic behavior: Since cyclic granularities are considered instead of linear granularities, clustering would group keys that have similar behavior across these cyclic granularities. This implies they will be grouped according to their periodic behavior and not on the linear stretch of time over which they are observed. -->

<!-- The following contributions are made through the following chapter: -->

<!--  * Present a cluster analysis of SGSC dataset to group households with similar periodic behavior -->
<!--  * Cluster validation by relating to external data -->

<!-- Electricity smart meter technology is increasingly being deployed in residential and commercial buildings. For example, in the state of Victoria, Australia, it is a state government policy that all households have a smart meter fitted (Victoria State Government, 2015), which has resulted in 97% of the approximately 2.1 million households in Victoria having a smart meter installed in their home. These smart meters collect energy usage information at half-hourly intervals, resulting in over 35 billion half-hourly observations per year across all households in the state. Clustering electricity consumption patterns can enable electricity utilities to develop targeted tariffs for individual groups alleviating the problem of volatility in production by capitalizing on the flexibility of consumers. -->

<!-- Clustering households using only smart meter consumption data could also provide value in a societal setting by combining the findings and external data like weather conditions, socio-economic or other demographic factors of those households. -->

<!-- Background literature and shortcomings -->

<!-- # Preliminary exploratoration -->

<!-- ## Electricity demand data -->

<!-- _Electricity use interval reading data_   -->
<!-- The data from [SGSC consumer trial data](https://data.gov.au/data/dataset/smart-grid-smart-city-customer-trial-data) is available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee). -->

<!-- ### Data source -->

<!-- The entire data is procured from CSIRO. A subset of this data is also available from [SGSC consumer trial data](https://data.gov.au/data/dataset/smart-grid-smart-city-customer-trial-data) is available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee). It consists of the following data sets.  -->
<!-- _1. CustomerData:_ 78720 customers with 62 variables about them  -->
<!-- _2. EUDMData:_ 300 billion half-hourly consumption level data   -->
<!-- _3. OffersData:_ Method of contact to customer to join SGSC customer trial, either door-to-door (D2D) or via Telesales   -->
<!-- _4. PEResponseData:_ Peak Events response customer wise   -->
<!-- _5. PETimesData:_ Peak Events time stamps   -->

<!-- Only _CustomerData_ and _EUDMData_ are relevant for the clustering goals of this paper. _EUDMData_ contains half-hourly general supply in KwH for 13,735 customers, resulting in 344,518,791 observations in total. `CustomerData` provides demographic data for 78,720 customers with information about their Local Government Area amongst others. -->

<!-- ## LGA and weather data -->

<!-- Since the smart meters have been installed at different dates for each household, it is reasonable to assume that the records are obtainable for different time lengths -->
<!-- for each household. Since, general supply is available for only 13,735 customers, we will restrict ourselves to look at the LGA information for these customers only. We find that there are only 26 LGA that is covered for these customers. -->

<!-- _Weather data_   -->
<!-- This data is obtained through [Australian Government Bureau of Meteorology](BOM)(http://www.bom.gov.au/) and provides hourly data for nearest weather stations for all the LGAs -->

<!-- This section familiarizes the 13,000 SGSC households through visualization and provide a detailed layout of data structures (like missing observations/number of customers/number of observations) and also the external data that needs to be utilized for validating the clustering process. The [ABS TableBuilder](https://www.abs.gov.au/websitedbs/censushome.nsf/home/tablebuilder) has census data from 2011 and 2016. The data is at SA2 and LGA levels. However, some of the LGA in NSW changed between 2011 and 2016 and hence there would not be a one-to-one correspondence between the LGAs. Weather, notably temperature (and humidity) can be the main driver(s) for energy usage. In NSW many households have electric heaters so their use can impact winter energy use and air-conditioners can impact summer energy use. Relevant weather data could be obtained from the [Bureau of Meteorology](www.bom.gov.au). Some weather stations have 30 minute (sometimes even smaller interval) weather data. Potentially, there could be lag effects of weather on energy usage which should be considered. -->

The remainder of the paper is organized as follows: Section&nbsp;\ref{sec:methodology} provides the clustering methodology introducing the features and distance metrics.
Section&nbsp;\ref{sec:validation} shows data designs to validate our methods and draw comparisons against several methods. Section&nbsp;\ref{sec:application} discusses the application of the method to a subset of the real data. Finally, we summarize our results and discuss possible future directions in Section&nbsp;\ref{sec:discussion}.

# Clustering methodology {#sec:methodology}

<!-- In contrast to models, a feature-based strategy is used to explicitly define or automatically extract the curves’ key time features, for instance by application of PCA on the daily curves [ -->

<!-- Most papers discussed in Tureczek2017-pb fail to accept smart meter readings as time series data, a data type which contains a temporal component. The omission of the essential time series features in the analysis leads to the application of methods that are not designed for handling temporal components. K-Means ignores autocorrelation, unless the input data is pre-processed. The clusters identified in the papers are validated by a variety of indices, with the most prevalent -->
<!-- being the cluster dispersion index (CDI) [22–24], the Davies–Bouldin index (DBI) [25,26] and the mean index adequacy (MIA) [8,13]. -->

The data set solely contains readings from smart meters and no information about the consumers' specific physical, geographical, or behavioural attributes. As a result, no attempt is made to explain why consumption varies. Instead, this work investigates how much energy usage heterogeneity can be found in smart meter data and what some of the most common electricity use patterns are. It is worth noting that when studying these dynamics, a variety of objectives may be pursued. One objective could be to group consumers with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. Most clustering algorithms offer only daily energy profiles throughout the hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar shapes across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification.

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on time-series features. First, we study the underlying distributions that may have resulted in different patterns across temporal granularities in order to identify a mechanism to classify them based on the similarity of those distributions. Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. This section presents the work flow of the methodology:

- _Data preparation_

@wang2020tsibble introduced the tidy "tsibble" data structure to support exploration and modeling of temporal data. A tsibble comprises an index, optional key(s), and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. For each key variable, the raw smart meter data is a sequence that is indexed by time and comprises values of several measurement variables at each time point. This sequence, though, could be depicted in a variety of ways. A shuffling of the raw sequence could reflect the distribution of hourly consumption over a single day, while another could indicate consumption over a week or a year. These temporal deconstructions of a time period into units such as hour-of-day, work-day/weekend are called cyclic temporal granularities. All cyclic granularities can be expressed in terms of the index set and could be augmented with the initial tsibble structure (index, key, measurements). It is worthwhile to note that the data structure changes while transporting from linear to cyclic scale of time as multiple observations of the measured variable would correspond to each category of the cyclic granularities. In this paper, quantiles are chosen to characterize the distributions for each category of the cyclic granularity. So, each category of a cyclic granularity corresponds to a list of numbers which is essentially few chosen quantiles of the multiple observations.

- _Finding significant cyclic granularities or harmonies_

\noindent These cyclic granularities are useful for exploring repetitive patterns in time series data that get lost in the linear representation of time. It is advantageous to consider only those cyclic granularities across which there is a significant repetitive pattern for the majority of customers or noteworthy in an electricity-behavior context. In that case, when the customers are grouped, we can expect to observe some interesting patterns across the categories of the cyclic granularities considered. [XXX reference 2nd chapter] proposes a way to select significant cyclic granularities and harmonies which is used for this paper.

- _Individual or combined categories of cyclic granularities as DGP_

The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. In our work, we are using $i$ as denoting a customer and the underlying variable as the electricity demand. So $f_i(v)$ is the distribution of household $i$ and $v$ is electricity demand. In this work, instead of considering the probability distributions of the linear time series, we assume that the measured variables across different categories of any cyclic granularity are from data generating processes.
Hence, we want to be able to cluster distributions of the form $f_{i,A,B \dots, {N_C}}(v)$, where $A, B$ represent the cyclic granularities under consideration such that $A = \{a_j: j=1, 2, \dots J\}$,  $B = \{b_k: k  = 1, 2, \dots K\}$ and so on. We consider individual each category of a cyclic granularity ($A$) or combination of categories for interaction of cyclic granularities (for e.g. $A*B$) to have a distribution. For example, let us consider we have two cyclic  granularities of interest, $A = {0, 1, 2, \dots, 23}$ representing hour-of-day and  $B = \{Mon, Tue, Wed, \dots, Sun\}$ representing day-of-week. 
each customer $i$ consist of a collection of probability distributions. In case individual granularities ($A$ or $B$) are considered there are  $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction,  $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. As a result, a distance between collections of these univariate probability distributions is required. Depending on the objective of the problem, there could be many approaches to considering such distances. This paper considers two approaches, which are explained in the next segment.

- _Distance metrics_

Considering each individual or combined categories of cyclic granularities as a data generating process lead to a collection of conditional distributions for each customer $i$. The (dis) similarity between each pair of observations should be obtained by combining the distances between these collections of conditional distributions such that the resulting metric is a distance metric, which could be fed into the clustering algorithm. Two types of distance metric is considered:

<!-- The choice of distance measures is a critical step in clustering. It defines how the similarity of two elements (x, y) is calculated. -->

**Inter-category distances**

This distance matrix considers two objects to be similar if every category of an individual cyclic granularity or combination of categories for interacting cyclic granularities have similar distributions. In this study, the distribution for each category is characterized using deciles and the distances between distributions are computed by using the Jensen-Shannon distance, which is symmetric and hence could be used as a distance measure.

\noindent The total distance between two elements $x$ and $y$ is then defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity A) or  $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$. When combining distances from individual cyclic granularities $A$ and $B$, $$S^{A, B}_{x, y} = S^A_{x,y}/J + S^B_{x,y}/K$$ is used, which could also be shown to be a distance metric easily. This is shown for cyclic granularity $A$ and $B$, but could be practically extended to more granularities.

**Intra-category distances**

Compute weighted pairwise distances ($wpd$) (XXX reference) for all considered granularities for all objects. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction and is estimated by the maximum pairwise distances between consecutive between consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting pattern is expected, whereas a lower value would indicate otherwise.

\noindent Distance between objects is then taken as the euclidean distances between them with the granularities being the variables and $wpd$ being the value under each variable. Since Euclidean distance is chosen, the observations with high values of features ($wpd$ values) will be clustered together. The same holds true for observations with low values of features. Thus this distance matrix would be useful to group customers that have similar significance of patterns across different granularities.

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of households showing similar periodic behavior by conidering $wpd$ vlaues for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

<!-- grouping probability distributions across a harmony. This clustering algorithm is adopted to remove or appropriately adjust for auto correlation and unequal length in the data. The method could be further extended by clustering probability distributions conditional on one or more cyclic granularities. The following are some of the advantages of our proposed method. -->

- _Pre-processing steps_

    **Handling trend, seasonality, non-stationarity and auto-correlation:** Trend and seasonality are fundamental characteristics of time series data, and it is reasonable to define a time series according to its degree of trend and seasonality. These characteristics of the time series are lost or handled independently by considering probability distributions ( trend is lost) across categories of cyclic granularities (by independently modeling all seasonal fluctuations), and so there is no need to de-trend or de-seasonalize the data before conducting the clustering method. There is no need to omit holiday or weekend patterns for similar reasons.

    **Data transformation:** Robust scaling method is used before computing the Inter-category distances and NQT is built-in transformation used for computation of $wpd$, which forms the basis of Intra-category distances.

- _Clustering algorithm_

In the analysis of energy smart metre data, K-Means or hierarchical clustering are often employed. These are simple and effective techniques that work well in a range of scenarios. For clustering, both employ a distance measure, and the distance measure chosen has a major influence on the structure of the clusters. We employ agglomerative hierarchical clustering in conjunction with Ward's criteria (XXX reference). Individual entities with the highest similarity computed using the desired distance metrics are sequentially merged using agglomerative algorithms. We can possibly employ any clustering method that supports the given distance metric as input.

- _Characterization of clusters_

Depending on the distance measure utilized for the study, the cluster characterization technique will differ. Clusters that utilise intra-category distances are characterised using multi-dimensional scaling and parallel coordinate displays. For inter-category distances, the distribution across major granularities may be presented to ensure that the goal of similar shapes within clusters and distinct shapes across clusters is met. This technique may potentially make advantage of multi-dimensional scaling.

\noindent Multidimensional scaling (MDS) (XXX reference) refers to a family of methods that analyse a matrix of distances or dissimilarities to provide a representation of the data points in a reduced-dimension space. There are many kinds of MDS, but they all solve the same fundamental issue: Given a $n*n$ matrix of dissimilarities and a distance measure, identify a configuration of $n$ points $x_1, x_2, \dots, x_n$ in the reduced dimension space $R^q$ ($q<p$) where the distance between the points is near to the dissimilarity between the points. All techniques must determine the coordinates of the points as well as the space dimension, $q$. Metric and nonmetric MDS are the two main kinds of MDS. Metric MDS methods presume a functional connection between the interpoint distances and the supplied dissimilarities and assume that the data are quantitative. We use metric MDS.

\noindent Parallel coordinate plots (XXX reference) Parallel coordinates have been extensively used to display high-dimensional and multivariate data, allowing for the detection of patterns within the data via visual grouping.

<!-- - A random sample  of the original data is taken for clustering analysis and includes missing and noisy observations (detailed description in Appendix) -->

<!-- - All harmonies are computed for each customer in the sample. -->
<!-- Cyclic granularities which are clashes for all customers in the sample are removed. -->

<!-- - It is worth noting that a number of other solutions may be considered at the pre-processing stage of the method. We have considered a) Normal-Quantile Transform and b) Robust transformation. -->

<!-- - Two methods are considered for computing dissimilarity between two customers. The first one involves computing according to one granularity is computed as the sum of the JS distances between distribution of all the categories of the granularity. When we consider more than one granularity, we consider the sum of the  average distances for all the granularity so that the combined metric is also a distance. -->

<!-- - Given the scale of dissimilarity among the energy readings, the model chooses optimal number of clusters -->

<!-- - Once clusters have been allocated, the groups are explored visually. -->

<!-- - Results are reported and compared. -->

<!-- Two methods are used for computing distances between subjects and then hierarchical clustering algorithm is used. -->

<!-- The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the probability distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. In our work, we are using $i$ as denoting a customer and the underlying variable as the electricity demand. So $f_i(v)$ is the distribution of household $i$ and $v$ is electricity demand. -->

<!-- We want to cluster distributions of the form $f_{i,j,k}(v)$, where $i$ and $j$ denote  -->

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of households showing similar periodic behavior by conidering $wpd$ vlaues for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

<!-- ### Notations -->

>>>>

theme_characterisation <- function() {

  theme_bw() + 
      theme(strip.text = element_text(size = 7, margin = margin(b = 0, t = 0))) +
   theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  theme(panel.spacing =unit(0, "lines")) + 
    theme(axis.text.x = element_text(angle=90, hjust=1, size = 7)) +
    theme(legend.position = "bottom")+
    theme(strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm")) +
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "grey"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "grey"))+
  theme(legend.position = "bottom")
}
```
